Ethnicity sensitive author disambiguation of semi-labeled corpora of scientific articles

### 2015-08-24 paper/paper.tex
save

Author name disambiguation in bibliographic databases is the problem of
grouping together scientific publications written by a same person, accounting
for potential homonyms and/or synonyms. Among solutions to this problem,
digital libraries are increasingly offering tools for authors to manually
curate their publications and claim which ones are theirs. Indirectly, these
tools allow for the inexpensive collection of large annotated training data,
which can be further leveraged to build a complementary automated
disambiguation system capable of inferring patterns for identifying
publications written by a same person.  Building upon more than 1 million of
publicly released crowdsourced annotations, we propose an automated author
disambiguation solution exploiting this data (i) to learn an accurate
classifier for identifying corefering authors and (ii) to guide the clustering
of scientific publications by distinct authors in a semi-supervised way. To the
best of our knowledge, our analysis is the first to be carried out on data of
this size and coverage. With respect to the state of the art, we validate the
general pipeline used in most existing solutions, and improve by proposing
phonetic-based blocking strategies, thereby increasing recall, and strong
ethnicity-sensitive features for learning a linkage function, thereby tailoring
disambiguation to non-Western author names whenever necessary.

@@

In academic digital libraries, author name disambiguation is the problem of
grouping together publications written by a same person.  Author name
disambiguation is often a difficult problem because an author may use different
spellings or name variants across her career (synonymy) and/or distinct authors may
share the same name (polysemy). Most notably, author disambiguation is often even more
troublesome in the case of researchers from non-Western cultures, where
personal names may be traditionally less diverse (leading to homonym issues) or
for which transliteration to Latin characters may not be unique (leading to
synonym issues). With the fast growth of the scientific literature, author
disambiguation has become a pressing issue because the accuracy of information
managed at the level of individuals directly affects the relevance search of
results (e.g., when querying for all publications written by a given author),
the reliability of bibliometrics and author rankings (e.g., citation counts or other impact
metrics, as studied in \citep{strotmann2012author}) or the relevance of scientific network analysis.
\glnote{Add more references.}

Efforts and solutions to author disambiguation have been proposed from various
communities \citep{liu2014author}. On the one hand, libraries have maintained
authorship control through manual curation, either in a centralized way by
hiring collaborators or by developing services inviting authors to register
their publications themselves (e.g., Google Scholar, Inspire-HEP, ...).
Recently, efforts also started to create persistent digital identifiers
assigned to researchers (e.g., ORCID, ResearchedID, ...), with the further
objective to embed these identifiers in the submission workflow of publishers
or repositories (e.g., Elsevier, arXiv, Inspire-HEP, ...), thereby univocally
solving any disambiguation issue. With the large cost of centralized manual
authorship control, or until the larger adoption of crowdsourced solutions, the
impact of these efforts are unfortunately limited by the efficiency, motivation
and integrity of their active contributors. Similarly, the success of
persistent digital identifier efforts is conditioned to a large and ubiquitous
adoption by both researchers and publishers. For these reasons, fully automated
machine learning-based methods have been proposed during the past decade to
provide immediate, less costly and efficient solutions to author
disambiguation. In this work, our goal is to explore and demonstrate how both
approaches can coexist and benefit from each other.  In particular, we study
how labeled data obtained through manual curation (either centralized or
crowdsourced) can be exploited (i) to learn an accurate classifier for
identifying corefering authors and (ii) to guide the clustering of scientific
publications by distinct authors in a semi-supervised way. Our analysis of parameters
and features on this large dataset reveals that the general pipeline
commonly used in existing solutions is an effective solution for author disambiguation.
In addition, and  as validated
by experimental results, we propose alternative strategies for blocking based on the
phonetization of author names, thereby increasing recall. We also propose
effective ethnicity-sensitive features for learning a linkage function,
thereby tailoring author
disambiguation to non-Western author names whenever necessary.

@@

Using Recursive Feature Elimination \cite{guyon2002gene}, we next evaluate the
usefulness of all 15 standard and 7 additional ethnicity features for learning
the linkage function. The analysis consists in using the baseline algorithm
first using all 22 features, to determine the least discriminative from feature
importances \citep{louppe2013understanding}, and then re-learn the baseline
using all but that one feature. That process is repeated recursively until
eventually only one feature remains. Results are presented in
Figure~\ref{fig:rfe} for one the three folds, starting from the far right with
the baseline and \textit{Second given name} being the least important feature,
and ending on the left with all features eliminated but \textit{Chinese}. As
the figure illustrates, the most important features are ethnic-based features
(\textit{Chinese}, \textit{Other Asian}, \textit{Black}) along with
\textit{Co-authors}, \textit{Affiliation} and \textit{Full name}. Adding the remaining
other features only brings marginal improvements, with \textit{Journal},
\textit{Abstract}, \textit{Collaborations}, \textit{References} \textit{Given
name initial} and \textit{Second given name} being almost insignificant.
Overall, these results highlight the added value of the proposed ehtnic-based
features. Their duality in modeling both the similarity between author names
and their origins make them very strong predictors for author disambiguation.
The results also corroborates results from \citep{ferreira2010effective}, who
found that the similarity between co-authors was a highly discriminative
feature. It indeed appears in 4th in our ranking and brings very significant
improvements as the figure illustrates. In case computational complexity is a concern, this analysis also
underlines the fact that decent performance can be achieved using only a very
small set of features, as also observed in
\citep{treeratpituk2009disambiguating} or \citep{levin2012citation}.

>>>>

Author name disambiguation in bibliographic databases is the problem of
grouping together scientific publications written by a same person, accounting
for potential homonyms and/or synonyms. Among solutions to this problem,
digital libraries are increasingly offering tools for authors to manually
curate their publications and claim which ones are theirs. Indirectly, these
tools allow for the inexpensive collection of large annotated training data,
which can be further leveraged to build a complementary automated
disambiguation system capable of inferring patterns for identifying
publications written by a same person.  Building upon more than 1 million of
publicly released crowdsourced annotations, we propose an automated author
disambiguation solution exploiting this data (i) to learn an accurate
classifier for identifying corefering authors and (ii) to guide the clustering
of scientific publications by distinct authors in a semi-supervised way. To the
best of our knowledge, our analysis is the first to be carried out on data of
this size and coverage. With respect to the state of the art, we validate the
general pipeline used in most existing solutions, and improve by proposing
phonetic-based blocking strategies, thereby increasing recall, along with strong
ethnicity-sensitive features for learning a linkage function, thereby tailoring
disambiguation to non-Western author names whenever necessary.

@@

In academic digital libraries, author name disambiguation is the problem of
grouping together publications written by a same person.  Author name
disambiguation is often a difficult problem because an author may use different
spellings or name variants across her career (synonymy) and/or distinct authors may
share the same name (polysemy). Most notably, author disambiguation is often even more
troublesome in the case of researchers from non-Western cultures, where
personal names may be traditionally less diverse (leading to homonym issues) or
for which transliteration to Latin characters may not be unique (leading to
synonym issues). With the fast growth of the scientific literature, author
disambiguation has become a pressing issue because the accuracy of information
managed at the level of individuals directly affects the relevance search of
results (e.g., when querying for all publications written by a given author),
the reliability of bibliometrics and author rankings (e.g., citation counts or other impact
metrics, as studied in \citep{strotmann2012author}) or the relevance of scientific network analysis
\citep{newman2001structure}.

Efforts and solutions to author disambiguation have been proposed from various
communities \citep{liu2014author}. On the one hand, libraries have maintained
authorship control through manual curation, either in a centralized way by
hiring collaborators or by developing services inviting authors to register
their publications themselves (e.g., Google Scholar, Inspire-HEP, ...).
Recently, efforts also started to create persistent digital identifiers
assigned to researchers (e.g., ORCID, ResearchedID, ...), with the further
objective to embed these identifiers in the submission workflow of publishers
or repositories (e.g., Elsevier, arXiv, Inspire-HEP, ...), thereby univocally
solving any disambiguation issue. With the large cost of centralized manual
authorship control, or until the larger adoption of crowdsourced solutions, the
impact of these efforts are unfortunately limited by the efficiency, motivation
and integrity of their active contributors. Similarly, the success of
persistent digital identifier efforts is conditioned to a large and ubiquitous
adoption by both researchers and publishers. For these reasons, fully automated
machine learning-based methods have been proposed during the past decade to
provide immediate, less costly and satisfactory solutions to author
disambiguation. In this work, our goal is to explore and demonstrate how both
approaches can coexist and benefit from each other.  In particular, we study
how labeled data obtained through manual curation (either centralized or
crowdsourced) can be exploited (i) to learn an accurate classifier for
identifying corefering authors and (ii) to guide the clustering of scientific
publications by distinct authors in a semi-supervised way. Our analysis of parameters
and features on this large dataset reveals that the general pipeline
commonly used in existing solutions is an effective approach for author disambiguation.
In addition, and  as validated
by experimental results, we propose alternative strategies for blocking based on the
phonetization of author names, thereby increasing recall. We also propose
effective ethnicity-sensitive features for learning a linkage function,
thereby tailoring author
disambiguation to non-Western author names whenever necessary.

@@

Using Recursive Feature Elimination \cite{guyon2002gene}, we next evaluate the
usefulness of all 15 standard and 7 additional ethnicity features for learning
the linkage function. The analysis consists in using the baseline algorithm
first using all 22 features, to determine the least discriminative from feature
importances \citep{louppe2013understanding}, and then re-learn the baseline
using all but that one feature. That process is repeated recursively until
eventually only one feature remains. Results are presented in
Figure~\ref{fig:rfe} for one the three folds, starting from the far right with
the baseline and \textit{Second given name} being the least important feature,
and ending on the left with all features eliminated but \textit{Chinese}. As
the figure illustrates, the most important features are ethnic-based features
(\textit{Chinese}, \textit{Other Asian}, \textit{Black}) along with
\textit{Co-authors}, \textit{Affiliation} and \textit{Full name}. Adding the remaining
other features only brings marginal improvements, with \textit{Journal},
\textit{Abstract}, \textit{Collaborations}, \textit{References} \textit{Given
name initial} and \textit{Second given name} being almost insignificant.
Overall, these results highlight the added value of the proposed ehtnic-based
features. Their duality in modeling both the similarity between author names
and their origins make them very strong predictors for author disambiguation.
The results also corroborates results from \citep{kang2009co} or \citep{ferreira2010effective}, who
found that the similarity between co-authors was a highly discriminative
feature. It indeed appears in 4th in our ranking and brings very significant
improvements as the figure illustrates. In case computational complexity is a concern, this analysis also
underlines the fact that decent performance can be achieved using only a very
small set of features, as also observed in
\citep{treeratpituk2009disambiguating} or \citep{levin2012citation}.
